Reconstructing Propositional Proofs in Type Theory

### 2017-10-23 paper/sections/proof-reconstruction/resolve.tex
[ paper ] lots of changes.

First, we define \rm{rassoc}$_{⬓}$ function in~(\ref{def:right-square})
to convert a disjunctions or a conjunction into its right-associative form.
We use $\rm{rassoc}_{∧}$ to convert conjunctions and $\rm{rassoc}_{∨}$ for
disjunctions.

\begin{equation}
\label{def:right-square}
  \begin{aligned}
    &\hspace{.495mm}\rm{rassoc}_{⬓} : \Prop \to \Prop\\
    &\begin{array}{lll}
    \rm{rassoc}_{⬓} &((φ₁ ⬓ φ₂) ⬓ φ₃) &= \rm{rassoc}_{⬓}~(φ₁~⬓~(φ₂ ⬓ φ₃))\\
    \rm{rassoc}_{⬓} &(φ₁ ⬓ φ₂)        &= φ₁~∨~\rm{rassoc}_{⬓}~φ₂\\
    \rm{rassoc}_{⬓} &φ                &= φ
    \end{array}
  \end{aligned}
\end{equation}

\begin{example}\hspace{5mm}\\
\begin{enumerate}
\item $\rm{rassoc}_{∨}~(((p ∨ ((q ∨ r) ∨ s)) ∨ t) = p ∨ (q ∨ (r ∨ (s ∨ t)))$.
\item $\rm{rassoc}_{∧}~((((p ∧ q) ∧ r) ∧ s) ∧ t) = p ∧ (q ∧ (r ∧ (s ∧ t)))$.
\end{enumerate}
\end{example}

\begin{remark}
In \TPTP syntax, the formulas are in left-associative form by default.
Despite of that convention, \Metis understands a formula to be in right-associative form by default. As we mentioned above, this is a matter to take into account for
the proof-reconstruction. We opted to work as far as possible with right-associative formulas in~\cite{AgdaMetis}.
\end{remark}

\begin{mainlemma}
\label{lem:rassoc}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rassoc}_{⬓}~φ$.
\end{mainlemma}

\begin{equation*}
 \label{eq:build-or}
  \begin{aligned}
  &\hspace{.495mm}\rm{build}_{∨} : \Prop → \Prop → \Prop\\
  &\begin{array}{llll}
    \rm{build}_{∨}&ψ~ψ &= ψ\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if } ψ_{1} ≡ \rm{build}_{∨}~φ~ψ_{1}\\
    \rm{build}_{∨}&φ~(ψ₁ ∨ ψ₂)&=ψ &\text{ if } ψ_{2} ≡\rm{build}_{∨}~φ~ψ_{2}\\
    \rm{build}_{∨}&φ~ψ&=φ&\text{ otherwise.}
   \end{array}
  \end{aligned}
\end{equation*}

\begin{equation}
  \label{eq:ndisj-or}
  \begin{aligned}
  &\hspace{.495mm}\rm{ndisj}_{∨} : \NProp \to \NProp\\
  &\begin{array}{lll}
    \rm{ndisj}_{∨} &(¬ φ₁ ∨ φ₂) &=
        \begin{cases}
         ⊤, &\text{ if } φ₁ ∈_{∨} φ₂\\
         ⊤, &\text{ if } \rm{ndisj}_{∨}~φ₂≡ ⊤\\
         ¬ φ₁ ∨ \rm{ndisj}_{∨}~φ₂, &\text{ otherwise.}
        \end{cases}\\

  \rm{ndisj}_{∨} &(φ₁ ∨ φ₂)&=
        \begin{cases}
         ⊤, &\,\,\,\,\,\,\text{ if } ¬φ₁ ∈_{∨} φ₂ \\
         ⊤, &\,\,\,\,\,\,\text{ if } \rm{ndisj}_{∨}(φ₂) ≡ ⊤\\
         φ₁ ∨ \rm{ndisj}_{∨}~φ₂, &\,\,\,\,\,\,\text{ otherwise.}
        \end{cases}\\
    \rm{ndisj}_{∨}&φ &=φ
    \end{array}
  \end{aligned}
\end{equation}

The $\rm{build}_{∨}$ function defined above intends to construct a
disjunction from another disjunction, specifically, this functions will
try to rearrange the disjuncts in the source formula to match with the
target disjunction formula, similar as we describe above.

\begin{mainlemma}
\label{lem:build-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{build}_{∨}~φ~ψ$.
\end{mainlemma}

@@

\begin{mainlemma}
\label{lem:factor}
  $Γ ⊢ φ$ then $Γ ⊢ \ffactor~φ$, where,
  \begin{equation}
  \label{eq:factor-definition}
   \begin{aligned}
   &\hspace{.495mm}\ffactor : \Prop → \Prop\\
    &\begin{array}{llll}
     \ffactor&(φ₁ ∨ φ₂) &=φ₁,  &\text{ if } φ₁ ≡ \ffactor~φ₂\\
     \ffactor&φ &= φ, &\text{ otherwise.}
  \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
We define the $\rm{sbuild}_{∨}$ function in~(\ref{eq:strong-build-or}).
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{sbuild}_{∨}~φ~ψ$, where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\rm{sbuild}_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{lll}
    \rm{sbuild}_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor(\rm{build}_{∨}~φ₁~ψ ∨ \rm{build}_{∨}~φ₂~ψ)\\
    \rm{build}_{∨}&φ~ψ &= \rm{build}_{∨}~φ~ψ
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

\begin{example}
Let see the main idea behind the lem-sbuild$_{∨}$ lemma.
We build the disjunction $((p ∨ q) ∨ r)$ from the disjunction
$(r ∨ (q ∨ p))$.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

  \end{bprooftree}
\end{equation*}
\end{example}

Notice that using $\rm{sbuild}_{∨}$ we can build not only a disjunction
with the same disjuncts of the source formula but also a complete different
formula by adding new disjuncts to the source formula.

\begin{mainlemma}
  \label{lem:reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}~φ~ψ~$, where,
  \begin{align*}
    \label{eq:reorder-or}
    \begin{split}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨}~φ~ψ~= \rm{sbuild}_{∨}~(rassoc_{∨}~φ)~ψ.
    \end{split}
  \end{align*}
\end{mainlemma}

@@

We define similar functions and state some theorems as we previously did
for disjunctions, but instead, we treat only with conjunctive normal forms.

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$, where, 
    \begin{align*}
      \begin{split}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                  &\text{and } ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$, where, 
  \begin{equation*}
\label{eq:conjunct-or-definition}
  \begin{split}
    &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
    &\fconjunct_{∨}~φ~ψ =
    \begin{cases}
      ψ, &\text{ if }φ ≡ ψ\\
      ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
      ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
        &\text{and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
      ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
      φ, &\text{ otherwise.}
    \end{cases}
  \end{split}
  \end{equation*}
\end{mainlemma}

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align*}
      \begin{split}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

The \resolve rule incorporates the resolution theorem showed in
Fig.~\ref{fig:metis-inferences} for the \TSTP derivations.
This rule posses three arguments: a positive literal $l$ to apply the
resolution theorem, and two derivations.
The positive literal $l$ must occur in
the formula from the first derivation and the
negative literal must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

% \begin{figure}
\label{fig:resolve-tstp-example}
\begin{verbatim}
cnf(refute_0_4, plain, (~ r | p | q),
    inference(canonicalize, [], [normalize_0_5])).
cnf(refute_0_5, plain, (p | q | r),
    inference(canonicalize, [], [normalize_0_6])).
cnf(refute_0_6, plain, (p | q),
    inference(resolve, [$cnf(r)], [refute_0_5, refute_0_4])).
\end{verbatim}
% \caption{Excerpt of a \Metis \TSTP example for the \resolve rule.}
% \end{figure}

To emulate this rule, we perform a sequence of rearrangements inside the
formulas involved to match with the pattern expected by the resolution theorem
described above. We begin by defining a convenient version of
the binary resolution in Def.~\ref{eq:rsol}.

% \begin{equation}
% \label{eq:rsol}
% \begin{aligned}
% &\rm{rsol} : \Prop \to \Prop\\
%       \begin{split}
%         &\rm{rsol} : \Prop \to \Prop\\
%         &\rm{rsol}~φ =
%         \begin{cases}
%           φ₂, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁, φ₂ ≡ φ₄\\
%           φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁\\
%           φ, &\text{ otherwise.}
%         \end{cases}
%       \end{split}
%   \end{align*}
% \end{definition}

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rsol}~φ$.
\end{mainlemma}

In Def.~\ref{def:resolve}, the proposed rule for $resolve$
has four arguments. The first three were described
above, the last one, ψ, represents the expected conclusion in the \Metis
derivation.

@@

\begin{example}COMPLETAR ESTE EJEMPLO AL FINAL
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ ¬ r ∨ p ∨ q$}
  \AxiomC{$Γ ⊢ p ∨ q ∨ r $}
  \RightLabel{Theorem \ref{thm:resolve}}
  \BinaryInfC{$Γ ⊢ \fresolve~(¬ r ∨ p ∨ q)~(p ∨ q ∨ r)~r~(p ∨ q)$}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \end{bprooftree}
\end{equation*}
\end{example}

>>>>

First, we define \fassoc$_{⬓}$ function in~(\ref{def:right-square})
to convert a disjunction or a conjunction into its right-associative form.
The square symbol can be the conjunction symbol or the disjunction symbol.
We use $\fassoc_{∧}$ to convert conjunctions and $\fassoc_{∨}$ for
disjunctions.

\begin{equation}
\label{def:right-square}
  \begin{aligned}
    &\hspace{.495mm}\fassoc_{⬓} : \Prop \to \Prop\\
    &\begin{array}{lll}
    \fassoc_{⬓} &((φ₁~⬓~φ₂)~⬓~φ₃) &= \fassoc_{⬓}~(φ₁~⬓~(φ₂~⬓~φ₃))\\
    \fassoc_{⬓} &(φ₁~⬓~φ₂)        &= φ₁~∨~\fassoc_{⬓}~φ₂\\
    \fassoc_{⬓} &φ                &= φ
    \end{array}
  \end{aligned}
\end{equation}

\begin{myexample}\hspace{10cm}
\begin{enumerate}
\item $\fassoc_{∨}~(((p ∨ ((q ∨ r) ∨ s)) ∨ t) = p ∨ (q ∨ (r ∨ (s ∨ t)))$.
\item $\fassoc_{∧}~((((p ∧ q) ∧ r) ∧ s) ∧ t) = p ∧ (q ∧ (r ∧ (s ∧ t)))$.
\end{enumerate}
\end{myexample}

\begin{mainlemma}
\label{lem:rassoc}
  If $Γ ⊢ φ$ then $Γ ⊢ \fassoc_{⬓}~φ$.
\end{mainlemma}

\begin{myremark}
In \TPTP syntax, the formulas are in left-associative form by default.
Despite of that convention, \Metis understands a formula to be in right-associative form by default. This is a matter to take into account for the proof-reconstruction.
\end{myremark}

The $\fbuild_{∨}$ function defined in~(\ref{eq:build-or})
intends to construct a disjunction from another disjunction, specifically,
this functions will try to rearrange the disjuncts in the source formula to
match with the target disjunction formula.

\begin{mainlemma}
\label{lem:build-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fbuild_{∨}~φ~ψ$, where,
\begin{equation*}
  \begin{split}
  \label{eq:build-or}
  &\fbuild_{∨} : \Prop → \Prop → \Prop\\
  &\fbuild_{∨}~φ~ψ~ =%\\
  % &\hspace{3mm}
  \begin{cases}
  ψ, &\text{ if } φ ≡ ψ\\
  ψ, &\text{ if } ψ ≡ ψ₁ ∨ ψ₂\text{, } ψᵢ ≡ \fbuild_{∨}~φ~ψᵢ\text{ for some }i=1, 2\\
  φ, &\text{ otherwise.}
  \end{cases}
  \end{split}
\end{equation*}
\end{mainlemma}

@@

\begin{mainlemma}
\label{lem:factor}
 If $Γ ⊢ φ$ then $Γ ⊢ \ffactor~φ$, where,
\begin{align}
\begin{split}
  \label{eq:factor-definition}
  &\ffactor : \Prop → \Prop\\
  &\ffactor~φ~ =
  \begin{cases}
    φ₁,  &\text{ if }φ ≡ φ₁ ∨ φ₂ \text{ and } φ₁ ≡ \ffactor~φ₂\\
    φ,   &\text{ otherwise.}
  \end{cases}
\end{split}
\end{align}
\end{mainlemma}

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
We define the $\fsbuild_{∨}$ function in~(\ref{eq:strong-build-or}).
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{mainlemma}
\label{lem:sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fsbuild_{∨}~φ~ψ$, where,
\begin{equation}
\label{eq:strong-build-or}
 \begin{aligned}
     &\hspace{.495mm}\fsbuild_{∨} : \Prop → \Prop → \Prop\\
    &\begin{array}{lll}
    \fsbuild_{∨}&(φ₁ ∨ φ₂)~ψ &=\ffactor~(\fbuild_{∨}~φ₁~ψ ∨ \fbuild_{∨}~φ₂~ψ)\\
    \fsbuild_{∨}&φ~ψ &= \fbuild_{∨}~φ~ψ
     \end{array}
\end{aligned}
\end{equation}
\end{mainlemma}

\begin{myexample}
We build the disjunction $(p ∨ q) ∨ r$ from the disjunction
$r ∨ (q ∨ p)$ based on the idea of the Lemma~\ref{lem:sbuild-or}.
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ q$}
  \RightLabel{∨-intro₂}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \RightLabel{∨-intro₁}
  \UnaryInfC{$Γ ⊢ (p ∨ q) ∨ r$}

@@

  \end{bprooftree}
\end{equation*}
\end{myexample}

\begin{myremark}
Notice that using $\fsbuild_{∨}$ we can build not only a disjunction
with the same disjuncts of the source formula but also a complete different
disjunction by adding new disjuncts to the source formula.
\end{myremark}

The following lemma tries to reorder nested disjunctions
by forcing the formula to be in right-associative form in order to
apply Lemma~\ref{lem:sbuild-or}.

\begin{mainlemma}
  \label{lem:reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}~φ~ψ~$, where,
  \begin{align}
    \begin{split}
    \label{eq:reorder-or}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨}~φ~ψ~= \fsbuild_{∨}~(\fassoc_{∨}~φ)~ψ.
    \end{split}
  \end{align}
\end{mainlemma}

@@

Now, we define function to work with nested conjunctions. This will
help us in the end of this section to reorder conjunctive normal forms.
For instance, the $\freorder_{∧}$ function in~(\ref{eq:reorder-and}),
reorders nested conjunctions.

\begin{mainlemma}
  \label{lem:reorder-and}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧}~φ~ψ$, where,
    \begin{align}
      \begin{split}
      \label{eq:reorder-and}
        &\freorder_{∧} : \Prop → \Prop \to \Prop\\
        &\freorder_{∧}~φ~ψ =
        \begin{cases}
          φ, &\text{ if }φ ≡ ψ\\
          ψ₁ ∧ ψ₂, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\text{, }ψ₁ ≡ \freorder_{∧}~φ~ψ₁ \\
                   &\text{ and } ψ₂ ≡ \freorder_{∧}~φ~ψ₂\\
          φ,       &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct~φ~ψ, &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align}
\end{mainlemma}

\begin{myexample}\hspace{10cm}
\begin{enumerate}
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(r ∧ q ∧ p)~=~(r ∧ q ∧ p).$
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(r ∧ r ∧ p)~=~(r ∧ q ∧ p).$
  \item $\freorder_{∧}~(p ∧ q ∧ r)~(k ∧ q ∧ p)~=~(p ∧ q ∧ r).$
  \item $\freorder_{∧}~((p ∨ q) ∧ r)~(r ∧ (q ∨ p))~=~((p ∨ q) ∧ r).$
\end{enumerate}
\end{myexample}

In the last example, Item~4, we could not build the conjunction
$r ∧ (q ∨ p)$ since $p ∨ q$ is not syntactical equal to $q ∨ p$.
We solve this issue in the Lemma~\ref{lem:reorder-and-or} by
using the $\fconjunct_{∨}$ function defined in~(\ref{eq:conjunct-or}).

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$, where,
  \begin{equation*}
    \begin{split}
    \label{eq:conjunct-or}
      &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
      &\fconjunct_{∨}~φ~ψ =
      \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
        ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
           &\text{ and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
        φ, &\text{ otherwise.}
      \end{cases}
    \end{split}
  \end{equation*}
\end{mainlemma}

Now, we are able to reorder conjunctive normal forms using
$\freorder_{∧∨}$ function defined in (\ref{eq:reorder-and-or})
that unifies the previous lemma showed above.

\begin{mainlemma}
  \label{lem:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}~φ~ψ$, where,
   \begin{align*}
      \begin{split}
      \label{eq:reorder-and-or}
      &\freorder_{∧∨} : \Prop \to \Prop \to \Prop\\
      &\freorder_{∧∨}~φ~ψ=
        \begin{cases}
          ψ, &\text{ if } φ ≡ ψ\\
          ψ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \freorder_{∧∨}~φ~ψ₁\\
             &\text{and } ψ₂ ≡ \freorder_{∧∨}~φ~ψ₂\\
          φ, &\text{ if } ψ ≡ ψ₁ ∧ ψ₂\\
          \fconjunct_{∨}~φ~ψ &\text{ otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{mainlemma}

Now, we are in a position to reconstruct \resolve rule. This is the
first application of the Lemma \ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem showed in
Fig.~\ref{fig:metis-inferences} in the \TSTP derivations.
This rule has three arguments: a positive literal $l$ to apply the
resolution theorem and two derivations.
The positive literal $l$ must occur in
the formula from the first derivation and the
negative literal $¬ l$ must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

\begin{myexample}\hspace{10cm}
\begin{verbatim}
cnf(ref4, (~ r | p | q), inf(canonicalize, [norm5])).
cnf(ref5, (p | q | r), inf(canonicalize, [norm6])).
cnf(ref6, (p | q), inf(resolve, [$cnf(r)], [ref5, ref4])).
\end{verbatim}
In the excerpt above, we apply resolution for the first two formulas,
$¬ r ∨ p ∨ q$ and $p ∨ q ∨ r$. The last line tells us the literal mentioned
for resolution is $r$. Syntactically speaking,
we can not derive neither the conclusion $p ∨ q$ in \verb!ref6! nor apply the resolution theorem with \verb!ref4! and \verb!ref5! since the formulas does not fit the pattern required.

If the scenario would have other like replacing \verb!ref5! by
\begin{verbatim}
cnf(ref5, (r | p | q), inf(canonicalize, [norm6])).
\end{verbatim}
We have could derive $(p ∨ q) ∨ (p ∨ q)$ but again, that is not the expected result.
\end{myexample}

Therefore, we perform a sequence of rearrangements inside the
formulas involved to match with the pattern expected by the resolution theorem
described in Fig.~\ref{fig:metis-inferences}. Using reordering after
applying a convenient version of the resolution theorem defined
in~(\ref{eq:rsol}) we get the expect result.

\begin{mainlemma}
  \label{lem:rsol}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rsol}~φ$, where,
  \begin{equation}
    \label{eq:rsol}
    \begin{aligned}
    &\rm{rsol} : \Prop \to \Prop\\
    &\begin{array}{ll}
      \rm{rsol}~φ &=
        \begin{cases}
          φ₂, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₂)\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (¬φ₁ ∨ φ₄)\\\
          φ, &\text{ otherwise.}
        \end{cases}
      \end{array}
      \end{aligned}
\end{equation}
\end{mainlemma}

@@

\begin{myexample}
Continuing with the problem presented in the example above,
we can use Theorem \ref{thm:resolve} to derive $Γ ⊢ p ∨ q$.
Let be $φ₁ ≡ p ∨ q ∨ r$ and $φ₂ ≡ ¬ r ∨ p ∨ q$.

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ p ∨ q ∨ r $}
  \AxiomC{$Γ ⊢ ¬ r ∨ p ∨ q$}
  \RightLabel{Theorem \ref{thm:resolve}}
  \BinaryInfC{$Γ ⊢ \fresolve~(p ∨ q ∨ r)~(¬ r ∨ p ∨ q)~r~(p ∨ q)$}
  \RightLabel{by (\ref{eq:resolve})}
  \UnaryInfC{$Γ ⊢ \rm{rsol}~(\freorder_{∨}~φ₁~(r ∨ (p ∨ q))~∧~\freorder_{∨}~φ₂~(¬ r ∨ (p ∨ q))$}
  \RightLabel{by (\ref{eq:reorder-or})}
  \UnaryInfC{$Γ ⊢ \rm{rsol}~((r ∨ (p ∨ q)) ∧ (¬ r ∨ (p ∨ q)))$}
  \RightLabel{by (\ref{eq:rsol})}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \end{bprooftree}
\end{equation*}
\end{myexample}
